---
title: "Differences in the Human Gut Microbiome in Children with Autism Spectrum Disorder"
output: 
  html_document: 
    keep_md: yes
---

(This space likely for notes/explaining what the project is about)

### Loading the libraries
```{r}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
library(ggvenn)
library(VennDiagram)
library(shiny)
library(shinydashboard)
library(ggplot2)
library(purrr)
library(dplyr)
```

### Loading the .csv files
```{r}
ASD_meta_adundance <- read_csv(here("Berlin/Copy_others/Ally/ASD_meta_abundance_2 copy.csv"))
microbiome_diversity <- read_csv(here("Berlin/Copy_others/Ally/microbiome_diversity copy.csv"))
asd_group_tidier <- read_csv(here("Berlin/Copy_others/Ally/asd_group_tidier copy.csv"))
td_group_tidier <- read_csv(here("Berlin/Copy_others/Ally/td_group_tidier copy.csv"))
demographics_and_abundance <- read_csv(here("Berlin/Copy_others/Ally/demographics_and_abundance copy.csv"))
demo_venn <- read_csv(here("Berlin/Copy_others/Ally/demo_venn copy.csv"))
asd_B2 <- read_csv(here("Berlin/Copy_others/Ally/asd_B2 copy.csv"))
td_A2 <- read_csv(here("Berlin/Copy_others/Ally/td_A2 copy.csv"))
asd2 <- read_csv(here("Berlin/Copy_others/Ally/asd2 copy.csv"))
demographics2 <- read_csv(here("Berlin/Copy_others/Ally/demographics2 copy.csv"))
microbiome_big_diff_long <- read_csv(here("Berlin/Copy_others/Claire/microbiome_big_diff_long copy.csv"))
microbiome_avg_long <- read_csv(here("Berlin/Copy_others/Claire/microbiome_avg_long copy.csv"))
big_demographics_A <- read_csv(here("Berlin/Copy_others/Claire/big_demographics_A copy.csv"))
big_demographics_B <- read_csv(here("Berlin/Copy_others/Claire/big_demographics_B copy.csv"))
```
# Shiny


## Graphs comparing Genus, Species, and Unique Counts (Unique counts not coming up for some reason)
```{r, eval=F, echo=T}
ui <- fluidPage(    
  
  titlePanel("Abundance Comparison"),
  
  sidebarLayout(      
    
    sidebarPanel(
      selectInput("x", "Select X Variable", choices = c("group"), selected = "group"),
      selectInput("y", "Select Y Variable", choices = c("n_genus", "n_species", "unique_counts"), selected = "n_genus"),
      
      hr(),
      helpText("Reference: Zhou Dan, et al. (2020) Altered gut microbial profile is associated with abnormal metabolism activity of Autism Spectrum Disorder, Gut Microbes, 11:5, 1246-1267, DOI: 10.1080/19490976.2020.1747329
")
    ),
    
    mainPanel(
      plotOutput("Plot")  
    )
    
  )
)

server <- function(input, output) {
  
  output$Plot <- renderPlot({
    
    ggplot(microbiome_diversity, aes_string(x = input$x, y = input$y, fill= "group")) + geom_col(position = "dodge") + theme(axis.text.x = element_text()) + theme(plot.title = element_text(size = rel(1.5)))+ scale_fill_brewer(palette = "Set1")
  })
  }

shinyApp(ui, server)
```

## Graphs of Abundance Versus Demographics
```{r, eval=F, echo=T}
ui <- dashboardPage(
  dashboardHeader(title = "Abundance Versus Demographics"),
  dashboardSidebar(),
  dashboardBody(
  selectInput("x", "Select X Variable", choices = c("age", "gender"), selected = "age"),
  selectInput("y", "Select Y Variable", choices = c("abundance"), selected = "abundance"),
  plotOutput("plot", width = "500px", height = "400px"))
)

server <- function(input, output, session) { 
  output$plot <- renderPlot({
  ggplot(demographics_and_abundance, aes_string(x = input$x, y = input$y, fill= as.factor("abundance"))) + geom_col(position = "dodge") + theme(axis.text.x = element_text()) + theme(plot.title = element_text(size = rel(1.5)))+ scale_fill_brewer(palette = "Set1")
  })
  session$onSessionEnded(stopApp)
  }
shinyApp(ui, server)
```


# Venn Diagrams of Data


## Preparation of data for Venn Diagrams that compares genera composition between ASD and TD patients (Note: couldn't use a csv file to cut down chunks that were run because the files were too big)


### Preparing TD data:
```{r}
td_venn <- td_group_tidier %>%
  rename(genus_td = "genus") %>% 
  filter(abundance != 0) %>% 
  select(genus_td, patient_id)
```

### Preparing ASD data:
```{r}
asd_venn <-
  asd_group_tidier %>% 
  rename(genus_asd = "genus") %>% 
  filter(abundance != 0)
```

### Joining TD and ASD data:
```{r}
both_venn <- full_join(td_venn, asd_venn, by = "patient_id")
both_venn
```

```{r}
both_venn$genus_td <- as.factor(both_venn$genus_td)
both_venn$genus_asd <- as.factor(both_venn$genus_asd)
```

```{r}
both_venn2 <- 
  both_venn %>% 
  filter(genus_td != "NA" | genus_asd != "NA")
```


### Building Venn Diagram that compares genera composition between ASD and TD patients
```{r}
venn_data <- list(
  "TD Genera" = both_venn2$genus_td,
  "ASD Genera" = both_venn2$genus_asd
)
```


## General Comparison of Genome Overlap between ASD and TD patients
```{r}
ggvenn(venn_data, fill_color = c("red", "blue"))
```
## General Comparison of Genome Overlap between ASD and TD patients to Scale
```{r}
grid.newpage()
draw.pairwise.venn(area1 = 1264,                      
                   area2 = 1101,
                   cross.area = 1068, 
                   fill = c("red", "blue"), 
                   category = c("TD Genera", "ASD Genera"))
```

### Preparation of data for Venn Diagrams that compares genera composition between gender
```{r}
demo_venn
```
### Dividing genera by gender
```{r}
demo_venn_f <- 
  demo_venn %>% 
  filter(gender == "Female") %>% 
  rename(genus_f = "genus")
demo_venn_f
```
```{r}
demo_venn_m <-
  demo_venn %>% 
  filter(gender == "Male") %>% 
  rename(genus_m = "genus")
demo_venn_m
```
```{r}
demo_venn_f$genus_f <- as.factor(demo_venn_f$genus_f)
demo_venn_m$genus_m <- as.factor(demo_venn_m$genus_m)
```

## General Comparison of Genome Overlap between Gender
```{r}
demo_venn_data <- list(
  "Male Genera" = demo_venn_m$genus_m,
  "Female Genera" = demo_venn_f$genus_f
)
```

```{r}
ggvenn(demo_venn_data, fill_color = c("red", "blue"))
```

## General Comparison of Genome Overlap between Gender to Scale
```{r}
grid.newpage()
draw.pairwise.venn(area1 = 652+643,                      
                   area2 = 644,
                   cross.area = 643, 
                   fill = c("red", "blue"), 
                   category = c("Male Genera", "Female Genera"))
```

### Preparation of data for Venn Diagrams that compares genera composition between age groups

```{r}
age_venn <- full_join(asd2, demographics2, by = "sample_id") %>% 
  select(-c(sample_id, abundance, gender)) %>% 
  mutate(age = case_when(age <= 7 ~ "younger",
                         age > 7 ~ "older"))
age_venn$genus <- as.factor(age_venn$genus)
age_venn_y <-
  age_venn %>% 
  filter(age == "younger") %>% 
  rename(genus_y = "genus")

age_venn_o <-
  age_venn %>% 
  filter(age == "older") %>% 
  rename(genus_o = "genus")
```

## General Comparison of Genome Overlap between age groups
```{r}
age_venn_data <- list(
  "Younger Genera" = age_venn_y$genus_y,
  "Older Genera" = age_venn_o$genus_o
)
```

```{r}
ggvenn(age_venn_data, fill_color = c("red", "blue"))
```
## General Comparison of Genome Overlap between age groups to Scale
```{r}
grid.newpage()
draw.pairwise.venn(area1 = 359+926,                      
                   area2 = 926+11,
                   cross.area = 926, 
                   fill = c("red", "blue"), 
                   category = c("Younger Genera", "Older Genera"))
```


## Building Venn Diagram that compares unique species composition between ASD and TD patients
```{r}
species_venn_data2 <- list(
  "TD Species" = td_A2$tax_td,
  "ASD Species" = asd_B2$tax_asd
)
```


## General Comparison of unique species composition between ASD and TD patients
```{r}
ggvenn(species_venn_data2, fill_color = c("red", "blue"))
```


## General Comparison of unique species composition between ASD and TD patients to Scale
```{r}
grid.newpage()
draw.pairwise.venn(area1 = (766+3591),                      
                   area2 = (172 + 3591),
                   cross.area = 3591, 
                   fill = c("red", "blue"), 
                   category = c("TD Species", "ASD Species"))
```


# Heatmaps

### Colors used:
```{r}
mycol <- c("deepskyblue4", "darkred")
mycoln <- c("darkred", "lightpink", "deepskyblue4")
```

## Heatmap of difference between “a” and “b” for each high variation genera
```{r}
microbiome_big_diff_long %>% 
  ggplot(aes(x = reorder(genus, avg_abundance) , y = condition))+
  geom_tile(aes(fill = avg_abundance, ))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Abundance of Highest Variation Genera", x = "Genus", y = "Condition", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```


## Heatmap of the 20 most abundant genera
```{r}
microbiome_avg_long %>%
  filter(avg_abundance > 114 | genus == "Odoribacter" | genus == "Parasutterella") %>% 
  ggplot(aes(x = reorder(genus, avg_abundance) , y = condition))+
  geom_tile(aes(fill = avg_abundance))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Most Abundant Genera", y = "Condition", x = "Genus", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```

## Age Heatmaps


## Heatmap of the 20 most abundant genera in group A
```{r}
big_demographics_A %>%
  filter(genus == "Hungatella" | genus == "Faecalibacterium" | genus == "Phascolarctobacterium" | genus == "Flavonifractor" | genus == "Butyricicoccus" | genus == "Subdoligranulum" | genus == "Bilophila" | genus == "Butyricimonas" | genus == "Holdemanella" | genus == "Oscillibacter" | genus == "Fusicatenibacter" | genus == "Intestinimonas" | genus == "Ruthenibacterium" | genus == "Paraprevotella" | genus == "Escherichia" | genus == "Roseburia"| genus == "Eisenbergiella"| genus == "Alistipes"| genus == "Eubacterium"| genus == "Odoribacter") %>%
  ggplot(aes(x = reorder(genus, avg_abundance) , y = factor(age)))+
  geom_tile(aes(fill = avg_abundance))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Most Abundant Genera in A Group", y = "Age", x = "Genus", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```


## Heatmap of the 20 most abundant genera in group B
```{r}
big_demographics_B %>%
  filter(genus == "Hungatella" | genus == "Faecalibacterium" | genus == "Phascolarctobacterium" | genus == "Flavonifractor" | genus == "Butyricicoccus" | genus == "Subdoligranulum" | genus == "Bilophila" | genus == "Butyricimonas" | genus == "Holdemanella" | genus == "Oscillibacter" | genus == "Fusicatenibacter" | genus == "Intestinimonas" | genus == "Ruthenibacterium" | genus == "Lachnoclostridium" | genus == "Escherichia" | genus == "Roseburia"| genus == "Eisenbergiella"| genus == "Alistipes"| genus == "Eubacterium"| genus == "Odoribacter") %>%
  ggplot(aes(x = reorder(genus, avg_abundance) , y = factor(age)))+
  geom_tile(aes(fill = avg_abundance))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Most Abundant Genera in B Group", y = "Age", x = "Genera", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```


