---
title: "Ally's Clean data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
``` 

## Load the Data
```{r message=FALSE, warning=FALSE}
asd <- readr::read_csv("ASD_meta_abundance_2.csv")
asd
```


## Tidying the Data
```{r}
asd_tidy <- asd %>% 
  pivot_longer(-Taxonomy, 
               names_to = "subject", 
               values_to = "abundance") %>% 
  separate(subject, into = c("group", "number"), sep = 1)
asd_tidy
```


## Remove unidentified bacteria
```{r}
asd_tidy2 <- asd_tidy %>% 
  filter(str_detect(Taxonomy, "Unclassified") == FALSE)
asd_tidy2
```

## Tidy up the Taxonomy column
```{r}
asd_tidy3 <- 
  asd_tidy2 %>% 
  separate(Taxonomy, into = c("genus", "species"), sep = ";")
```

## Tidy up the species and genus and number columns
```{r}
asd_tidy4 <-
  asd_tidy3 %>% 
  separate(genus, into = c("g", "genus"), sep = "__") %>% 
  separate(species, into = c("s", "species"), sep = "__") %>% 
  select( -g, -s) %>% 
  rename(id = "number")
asd_tidy4
```


## Make genus and species and id into factors
```{r}
asd_tidy4$genus <- as.factor(asd_tidy4$genus)
asd_tidy4$species <- as.factor(asd_tidy4$species)
asd_tidy4$id <- as.factor(asd_tidy4$id)
```


## Break it into an ASD and TD groups
```{r}
asd_group <-
  asd_tidy4 %>% 
  filter(group == "B")
asd_group
```

```{r}
td_group <-
  asd_tidy4 %>% 
  filter(group == "A")
td_group
```

## Preliminary Analysis of TD and ASD groups
```{r}
td_diversity <- td_group %>% 
  filter(abundance != 0) %>% 
  summarize(n_genus = n_distinct(genus), n_species = n_distinct(species))
td_diversity
```

```{r}
asd_diversity <-
asd_group %>% 
  filter(abundance != 0) %>% 
  summarize(n_genus = n_distinct(genus), n_species = n_distinct(species))
asd_diversity
```
<<<<<<< HEAD
Note: The ASD Group has greater diversity than the TD group. 

## Plot the diversity of the ASD vs TD groups
```{r}
group <- c("td", "asd")
microbiome_diversity <-
  full_join(td_diversity, asd_diversity) %>% 
  cbind(group) %>% 
  select(group, n_genus, n_species)
microbiome_diversity
```

Here is the genus diversity:
```{r}
microbiome_diversity %>% 
  ggplot(aes(x = group, y = n_genus, fill = group)) +
  geom_col() +
  theme_linedraw() +
  labs(title = "Number of Genera per Group",
       x = "Group",
       y = "Number of Genera") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5)) +
  scale_fill_brewer(palette = "Set1")
```

Here is the species diversity:
```{r}
microbiome_diversity %>% 
  ggplot(aes(x = group, y = n_species, fill = group)) +
  geom_col() +
  theme_linedraw() +
  labs(title = "Number of Species per Group",
       x = "Group",
       y = "Number of Species") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5)) +
  scale_fill_brewer(palette = "Set1")
```

## Exporting CSVs
```{r}
write.csv(asd_tidy4, file = "microbiome.csv", row.names = FALSE)
```

```{r}
write.csv(asd_group, file = "asd_group.csv", row.names = FALSE)
```

```{r}
write.csv(td_group, file = "td_group.csv", row.names = FALSE)
>>>>>>> c55958c48b46b82ba90ce51fed413e8c84dde90b
```

