---
title: "Microbiome HeatMap"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(here)
```

# ASD/TD Condition Heatmap
## Loading the Data
```{r}
microbiome_untidy <- read_csv("C:/Users/Claire Chapman/Desktop/BIS15L_Project_Group11/untidy_data/ASD meta abundance 2.csv")
```
```{r}
microbiome_untidy <- janitor::clean_names(microbiome_untidy)
microbiome_untidy
```

## Finding the highest variation genera
#### Pivot longer and separate "condition" letter from "patient ID" to ultimately get an "a" and a "b" column.Tidying taxonomy column
```{r}
microbiome_longer <- microbiome_untidy %>% 
  pivot_longer(-taxonomy, names_to = "condition", values_to = "abundance")%>% 
  filter(str_detect(taxonomy, "Unclassified") == FALSE) %>% 
  separate(condition, into = c("condition", "patient_id"), sep = 1) %>%
  separate(taxonomy, into = c("genus", "species"), sep = ";s__") %>% 
  separate(genus, into = c("trash", "genus"), sep = "__") %>% 
  separate(species, into = c("species", "strain"), sep = " sp. ") %>%
  separate(species, into = c("extra_genus", "species"), sep = " ") %>% 
  mutate(cultured = ifelse(extra_genus == "uncultured", FALSE, TRUE)) %>% 
  select(- extra_genus, - trash) %>% 
  mutate(species = ifelse(species == genus, NA, species))
microbiome_longer
```

```{r}
microbiome_avg_long <- microbiome_longer %>% 
  group_by(genus, condition) %>% 
  summarise(avg_abundance = mean(abundance, na.rm = T)) %>% 
  select(genus, condition, avg_abundance)
microbiome_avg_long
```


#### Get "a" and "b" columns with "abundance" observations
```{r}
microbiome_wider <- microbiome_avg_long %>% 
  pivot_wider(names_from = "condition", values_from = "avg_abundance")
microbiome_wider
```

#### Create average value for each "genus" in each "condition", mutate a "difference" column for each "genera"
```{r}
microbiome_difference <- microbiome_wider %>% 
  group_by(genus) %>% 
  summarise(A = mean(a, na.rm = T), B = mean(b, na.rm = T)) %>% 
  mutate(difference = abs(A-B)) %>% 
  arrange(desc(difference))
microbiome_difference
```

#### Create a new dataframe, **microbiome_big_diff**, with the top most variable genera for all heatmap analysis
```{r}
microbiome_big_diff <- microbiome_difference %>%
  top_n(20)
microbiome_big_diff
```
## Making the heatmap + extra graphs
#### Graphing the differences between "a" and "b" for each of the high variation genera
```{r}
microbiome_big_diff %>% 
  select(genus, difference) %>% 
  group_by(genus) %>% 
  summarise(avg_difference = mean(difference)) %>% 
  ggplot(aes(x = reorder(genus, avg_difference), y = avg_difference))+
  geom_col(fill = "lightpink3")+
  coord_flip()+
  theme_minimal()+
  labs(title = "Difference Between Conditions in Highest Variable Genera", y = "Average Difference in Abundance", x = "Genus")
```


#### Making a new dataframe of the top 50 high variation genera
Getting rid of the difference column and making a "condition" column again + tidy taxonomy like I did for the main CSV
```{r}
microbiome_big_diff_long <- microbiome_big_diff %>% 
  select(-difference) %>% 
  pivot_longer(c("A", "B"), names_to = "condition", values_to = "avg_abundance")
microbiome_big_diff_long
```


#### Graphing the abundances of the high variation genera
```{r}
mycol <- c("deepskyblue4", "darkred")
mycoln <- c("darkred", "lightpink", "deepskyblue4")
```


```{r}
microbiome_big_diff_long %>% 
  ggplot(aes(x = reorder(genus, avg_abundance), y = avg_abundance, fill = condition))+
  geom_col(position = "dodge")+
  coord_flip()+
  scale_fill_manual(values = mycol)+
  labs(title = "Abundance of Highest Variable Genera by Condition", y = "Average Abundance", x = "Genus", fill = "Condition")
  
```

#### Heatmap of difference between "a" and "b" for each high variation genera
```{r}
microbiome_big_diff_long %>% 
  ggplot(aes(x = reorder(genus, avg_abundance) , y = condition))+
  geom_tile(aes(fill = avg_abundance, ))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Abundance of Highest Variation Genera", x = "Genus", y = "Condition", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```
## Most Abundant Genera
#### Heatmap of the 20 most abundant genera
```{r}
microbiome_avg_long %>%
  filter(avg_abundance > 114 | genus == "Odoribacter" | genus == "Parasutterella") %>% 
  ggplot(aes(x = reorder(genus, avg_abundance) , y = condition))+
  geom_tile(aes(fill = avg_abundance))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Most Abundant Genera", y = "Condition", x = "Genus", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```
```{r}
microbiome_avg_long %>% 
  filter(avg_abundance > 114 | genus == "Odoribacter" | genus == "Parasutterella") %>%  
  ggplot(aes(x = reorder(genus, avg_abundance), y = avg_abundance, fill = condition))+
  geom_col(position = "dodge")+
  coord_flip()+
  scale_fill_manual(values = mycol)+
  labs(title = "Most Abundant Genera by Condition", y = "Average Abundance", x = "Genus", fill = "Condition")
```


# Age Heatmap
## Loading Data
```{r}
demographics <- readr::read_csv("C:/Users/Claire Chapman/Desktop/BIS15L_Project_Group11/tidy_data/demographics_and_abundance.csv")
```

```{r}
names(demographics)
```
#### Separate condition and ID number
```{r}
demographics_untidy <- demographics %>%
  separate(sample_id, into = c("condition", "patient_id"), sep = 1)
demographics_untidy
```

#### Finding abundance by age for group A
```{r}
big_demographics_A <- demographics_untidy %>%
  filter(condition == "A") %>% 
  group_by(genus, age) %>% 
  summarise(avg_abundance = mean(abundance, na.rm = T)) %>% 
  arrange(desc(avg_abundance))
big_demographics_A
```

#### Making the heatmap with the 20 top genera
```{r}
big_demographics_A %>%
  filter(genus == "Hungatella" | genus == "Faecalibacterium" | genus == "Phascolarctobacterium" | genus == "Flavonifractor" | genus == "Butyricicoccus" | genus == "Subdoligranulum" | genus == "Bilophila" | genus == "Butyricimonas" | genus == "Holdemanella" | genus == "Oscillibacter" | genus == "Fusicatenibacter" | genus == "Intestinimonas" | genus == "Ruthenibacterium" | genus == "Paraprevotella" | genus == "Escherichia" | genus == "Roseburia"| genus == "Eisenbergiella"| genus == "Alistipes"| genus == "Eubacterium"| genus == "Odoribacter") %>%
  ggplot(aes(x = reorder(genus, avg_abundance) , y = factor(age)))+
  geom_tile(aes(fill = avg_abundance))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Most Abundant Genera in A Group", y = "Age", x = "Genus", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```

#### Finding abundance by age for group B
```{r}
big_demographics_B <- demographics_untidy %>%
  filter(condition == "B") %>% 
  group_by(genus, age) %>% 
  summarise(avg_abundance = mean(abundance, na.rm = T)) %>% 
  arrange(desc(avg_abundance))
big_demographics_B
```

#### Making the heatmap with top 20 abundant genera

```{r}
big_demographics_B %>%
  filter(genus == "Hungatella" | genus == "Faecalibacterium" | genus == "Phascolarctobacterium" | genus == "Flavonifractor" | genus == "Butyricicoccus" | genus == "Subdoligranulum" | genus == "Bilophila" | genus == "Butyricimonas" | genus == "Holdemanella" | genus == "Oscillibacter" | genus == "Fusicatenibacter" | genus == "Intestinimonas" | genus == "Ruthenibacterium" | genus == "Lachnoclostridium" | genus == "Escherichia" | genus == "Roseburia"| genus == "Eisenbergiella"| genus == "Alistipes"| genus == "Eubacterium"| genus == "Odoribacter") %>%
  ggplot(aes(x = reorder(genus, avg_abundance) , y = factor(age)))+
  geom_tile(aes(fill = avg_abundance))+
  coord_flip()+
  theme_minimal()+
  labs(title = "Most Abundant Genera in B Group", y = "Age", x = "Genera", fill = "Average Abundance")+
  scale_fill_gradientn(colours = mycoln, trans = "log")
```


